---
title: "Linear Regression Preprocessing and Modeling"
author: "Ya Wang"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_depth: 2
    theme: flatly
    highlight: textmate
  github_document: 
    toc: true
    toc_depth: 2
    html_preview: false
  pdf_document: default
---

<!-- badges: start -->

[![R-CMD-check](https://github.com/yawang0725/Biostat625_hw3/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/yawang0725/Biostat625_hw3/actions/workflows/R-CMD-check.yaml)

<!-- badges: end -->

# Linear Regression Preprocessing and Modeling

# Introduction

This package offers researchers a straightforward and comprehensive solution for modeling linear regression data while generating detailed statistical reports. It streamlines the analysis process, ensuring both ease of use and accuracy, making it an ideal tool for statistical exploration and reporting in research workflows.

This document demonstrates a comprehensive workflow for: - Cleaning data - Fitting a linear regression model - Evaluating model performance - Generating a complete regression summary

### Built With

Built Using Languages and Libraries Listed Below \* [R](https://www.r-project.org/)

## Table of Contents

-   [Getting Started](#GettingStarted)

    -   [Install](#Install)

    -   [Usage](#Usage)

-   [Example](#Example)

-   [Contributing](#Contributing)

-   [License](#License)

## Getting Started

### Install {#install}

Using [hw3](https://www.npmjs.org/), just run the following command:

```{r}
# install the package "devtools" first with:
# install.packages("devtools")

# install the new package
devtools::install_github("yawang0725/Biostat625_hw3")

# build the vignette
devtools::install_github('yawang0725/Biostat625_hw3', build_vignettes =T)

# browse the vignette 
browseVignettes(package = "hw3")
```

### Usage {#usage}

#### 1. `clean_data()`

Cleans a dataset by performing the following operations:

-   **Handling missing values:** Options for omitting rows with missing data or replacing missing values with the mean or median of the respective column.
-   **Handling outliers:** Options for removing or replacing outliers based on the Interquartile Range (IQR).
-   **Scaling numeric data:** Option to scale the numeric columns to have zero mean and unit variance.
-   **Removing constant columns:** Removes columns with constant values, which do not contribute to the predictive power of the model.

```{r}
library(hw3)
cleaned_iris <- clean_data(iris, 
                           method_missing = "omit", 
                           method_outliers = "none", 
                           scale_data = TRUE, 
                           remove_constant = TRUE)
```

**Parameters:**

-   `dataset`: A `data.frame` to be cleaned.
-   `method_missing`: Method to handle missing data (`"omit"`, `"mean"`, `"median"`).
-   `method_outliers`: Method to handle outliers (`"remove"`, `"replace"`, `"none"`).
-   `scale_data`: Boolean to scale numeric data (default is `FALSE`).
-   `remove_constant`: Boolean to remove constant columns (default is `FALSE`).

**Return:**

-   A cleaned `data.frame`.

#### 2. `fit_linear_model()`

Fits a linear regression model using Ordinary Least Squares (OLS) and extracts relevant statistical measures such as estimated coefficients, residuals, and variance-covariance matrix.

```{r}
fit_result <- fit_linear_model(dataset = cleaned_iris, 
                        response_var = "Sepal.Length", 
                        predictor_vars = c("Sepal.Width", "Petal.Length", "Petal.Width"))
```

**Parameters:**

-   `dataset`: A `data.frame` containing the dataset.
-   `response_var`: The name of the response (dependent) variable.
-   `predictor_vars`: A vector of predictor (independent) variable names.
-   `intercept`: Boolean indicating whether to include an intercept term (default is `TRUE`).

**Return:**

-   A list containing:
-   `betahat`: Estimated regression coefficients.
-   `y_hat`: Predicted values.
-   `residuals`: Residuals of the model.
-   `var_betahat`: Variance-covariance matrix for the regression coefficients.
-   `se_betahat`: Standard errors of the regression coefficients.
-   `mse`: Mean squared error of the model.
-   `degree`: Degrees of freedom.

#### 3. `extract_stats()`

Extracts statistical measures (t-statistics, p-values, confidence intervals) from a linear regression model fit result.

```{r}
stats <- extract_stats(fit_result, alpha = 0.01)
```

**Parameters:**

-   `fit_result`: A list containing the results from a linear regression model.
-   `alpha`: Significance level for confidence intervals (default is `0.05`).

**Return:**

-   A list containing:
-   `t_statistic`: t-statistics for each coefficient.
-   `p_value`: p-values for each coefficient.
-   `conf_int`: Confidence intervals for each coefficient.
-   `mse`: Mean squared error.

#### 4. `evaluate_model_performance()`

Evaluates the performance of a linear regression model by computing R-squared, Adjusted R-squared, F-statistic, and its associated p-value.

```{r}
performance_metrics <- evaluate_model_performance(
  fit_result = fit_result,
  dataset = cleaned_iris,
  response_var = "Sepal.Length",
  predictor_vars = c("Sepal.Width", "Petal.Length", "Petal.Width"))
```

**Parameters:**

-   `fit_result`: A list containing the results from a linear regression model.
-   `dataset`: The dataset used to fit the model.
-   `response_var`: The name of the response variable.
-   `predictor_vars`: A vector of predictor variables.

**Return:**

-   A list containing:
-   `R_squared`: The R-squared value.
-   `Adj_R_squared`: The Adjusted R-squared value.
-   `F_statistic`: The F-statistic value.
-   `p_value_F`: The p-value for the F-statistic.

#### 5. `generate_full_summary()`

Generates a complete regression summary, including: - Regression equation - Coefficient table with estimates, standard errors, t-values, p-values, and confidence intervals - Model fit statistics (R-squared, Adjusted R-squared, F-statistic, and its p-value) - Model evaluation (e.g., how well the model fits the data)

```{r}
generate_full_summary(cleaned_iris, 
                      response_var = "Sepal.Length", 
                      predictor_vars = c("Sepal.Width", "Petal.Length", "Petal.Width"))
```

**Parameters:**

-   `dataset`: A `data.frame` containing the dataset.
-   `response_var`: The name of the response (dependent) variable.
-   `predictor_vars`: A vector of predictor (independent) variable names.
-   `intercept`: Boolean indicating whether to include an intercept term (default is `TRUE`).

**Return:**

-   A printed model summary with all relevant statistics.

## Example {#example}

Input dataset: `iris` from `library(datasets)` (`data(iris)`). We compared the results of fitting a linear regression model using the custom function `fit_linear_model()` and the base `lm()` function. To ensure correctness, we used `all.equal()` to verify that the model coefficients, t-values, and p-values from both methods were virtually identical, with coefficients matching exactly and only minor differences in t-values and p-values (relative differences of 0.0134 and 5.17e-06, respectively). For efficiency, we employed `bench::mark()` to benchmark the performance of both methods. The custom method was significantly faster than `lm()`, processing three times more iterations per second, with slightly lower memory usage (32.8KB vs. 47.6KB). The number of garbage collection events was similar for both methods, and the total runtime was 378ms for the custom method versus 433ms for `lm()`. These comparisons demonstrate that the custom method excels in both accuracy and computational efficiency while providing results consistent with the standard R function.

```{r}
 # Clean the dataset
cleaned_iris <- clean_data(iris, 
                           method_missing = "omit", 
                           method_outliers = "none", 
                           scale_data = TRUE, 
                           remove_constant = TRUE)

# Fit the model using our custom function
custom_fit <- fit_linear_model(dataset = cleaned_iris, 
                               response_var = "Sepal.Length", 
                               predictor_vars = c("Sepal.Width", "Petal.Length", "Petal.Width"))

# Fit the model using lm()
base_model <- lm(Sepal.Length ~ Sepal.Width + Petal.Length + Petal.Width, data = cleaned_iris)

# Extract summary statistics from lm()
base_summary <- summary(base_model)
base_coef <- base_summary$coefficients

# Extract statistics from custom implementation
cat("Comparing coefficients:\n")
custom_stats <- extract_stats(custom_fit)
coef_match <- all.equal(as.vector(custom_fit$betahat), unname(coef(base_model)))
print(coef_match)

# Compare t-values and p-values
cat("\nComparing t-values:\n")
t_match <- all.equal(as.vector(custom_stats$t_statistic), unname(base_coef[, "t value"]))
print(t_match)

cat("\nComparing p-values:\n")
p_match <- all.equal(as.vector(custom_stats$p_value), unname(base_coef[, "Pr(>|t|)"]))
print(p_match)

library(bench)

benchmark_result <- bench::mark(
  lm_result = lm(Sepal.Length ~ Sepal.Width + Petal.Length + Petal.Width, data = cleaned_iris),
  custom_result = fit_linear_model(
    dataset = cleaned_iris, 
    response_var = "Sepal.Length", 
    predictor_vars = c("Sepal.Width", "Petal.Length", "Petal.Width")
  ),
  check = FALSE  
)

print(benchmark_result)
```

## Contributing {#contributing}

Ya Wang

## License {#license}

This project is licensed under the **GNU General Public License v3.0 or later (GPL-3.0+)**. You are free to use, modify, and distribute this software under the terms of this license.

For more details, see the [LICENSE](./LICENSE) file or visit the [GNU GPL website](https://www.gnu.org/licenses/gpl-3.0.en.html).
